---
title: "01_integration_and_LC_selection"
output: html_document
date: "2026-02-03"
---
# overview
1. how do we decided which clusters are target LC cells based on the integrated results



```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 6,
  fig.height = 5
)

library(Seurat)
library(harmony)
library(ggplot2)
library(dplyr)

source("../code/func/integration_functions.R")
```

# Plotting

## load the data that is generated from 01 script (`./code/01_build_full_object.R`). 
```{r load-data}
combined <- readRDS("../scratch/LC_clustered_harmony_1.rds")
```

## Cluster structure after integration on UMAP
```{r}
DimPlot(combined, reduction = "umap", group.by = "seurat_clusters",
        label = TRUE, repel = TRUE, pt.size = 0.3)+ NoLegend()
DimPlot(combined, reduction = "umap", group.by = "batch", pt.size = 0.3)
```

## show marker genes expression for these clusters
the canonical LC/noradernergic markers 
```{r lc-markers, fig.width=10, fig.height=6}
lc_markers <- c("Dbh", "Slc6a2", "Th", "Slc18a2", "Tacr3")
FeaturePlot( combined, features = lc_markers, reduction = "umap", pt.size = 0.3, ncol = 3, raster=FALSE, 
             min.cutoff = "q5", max.cutoff = "q95", cols = c("lightgrey", "firebrick3"))
```


```{r lc-dotplot, fig.width=5, fig.height=10}
DotPlot(
  combined,scale = FALSE,
  features = lc_markers,
  group.by = "seurat_clusters"
) + RotatedAxis()

```

seems cluster 9 adn maybe 26/35 is the target. 

```{r}
# 1) compute average expression per cluster (defaults to assay="RNA", slot="data" so using log CP10k)
avg_list <- AverageExpression(
  combined,
  features = lc_markers[1:3],
  group.by = "seurat_clusters",
  slot = "data",
  verbose = FALSE
)

# avg expression matrix: 3x num_cluster
avg_mat <- avg_list$RNA

# 2) find clusters where all first-3 markers have mean expression > 2
keep_clusters <- colnames(avg_mat)[colSums(avg_mat > 1) == length(lc_markers[1:3])]
keep_clusters
```


```{r}
lc_candidate_clusters <-  as.integer(sub("^g", "", keep_clusters))

```


```{r}
DimPlot(
  combined,
  reduction = "umap",
  group.by   = "seurat_clusters",
  cells      = WhichCells(combined, idents = lc_candidate_clusters),
  label      = TRUE)
lc_cells <- WhichCells(combined, idents = lc_candidate_clusters)
```


```{r,fig.width=10, fig.height=6}
FeaturePlot(
  combined,
  features   = lc_markers,
  reduction  = "umap",
  cells      = lc_cells,
  pt.size    = 0.3,
  ncol       = 3,
  raster     = FALSE,
  min.cutoff = "q5",
  max.cutoff = "q95",
  cols       = c("lightgrey", "firebrick3")
)
```




## basd on this it seems that we are interested in 9 and 26
```{r}

VlnPlot(combined,
  features = lc_markers,
  idents = lc_candidate_clusters,
  pt.size = 0,ncol = 3)

```


never mind because 26 has too low of expressions. 

> We initially considered clusters 9 and 26 as LC candidates. Cluster 9 shows strong and consistent expression of Dbh, Slc6a2, Th, and Slc18a2, whereas cluster 26 has weak/incomplete marker expression. We therefore define LC as cluster 9 only.

> The LC cluster(s) used in downstream analyses and in code/02_define_LC_population.R were defined as cluster 9 only.

**Based on co-expression of these markers, we define LC as cluster 9.**

